---
title: "Time Usage in Europe"
author: "Sofia Milone"
format: html
editor: visual
---

# TIME USAGE IN EUROPE

Time is the most valuable and precious resource we have. But what people do with it? This is what we are trying to understand with the following analysis. The data used was provided by EUROSTAT. They are conducting national surveys (HETUS, Harmonised European time use surveys) in EU countries to understand how people spent their time. So far, there are only 2 rounds of survey. The first one called HETUS 2000, in which participated 15 countries, and the second one (HETUS 2010) , where the participating countries were 18.

To know about the investigations: [`Time use survey (tus) (europa.eu)`{=html}](https://ec.europa.eu/eurostat/cache/metadata/en/tus_esms.htm)

To get the data analysed in this notebook: [Statistics \| Eurostat (europa.eu)](https://ec.europa.eu/eurostat/databrowser/view/tus_00age__custom_9558191/default/table?lang=en&page=time:2000)

## The Dataset

In the initial dataset we can find 7 variables:

-   Frequency - Character

-   Sex - categorical with 3 categories: Female (F), Male (M) and Total (T)

-   Age - ordinal with different classes

-   Activities - There were selected 56 activities in total, divided in 8 macroareas :

    -   Personal care

    -   Eat

    -   Sleep

    -   Work

    -   Study

    -   Household and family care

    -   Leisure, social time

    -   Unspecified Use of time

-   Country

-   The time spent on the activity

-   The partecipation time= *"mean time spent in the activities by those individuals who took part in the activity "* (EUROSTAT)

-   Participation rate:= *"the proportion of the individuals that spent some time doing the activities"* (EUROSTAT)

```{r, warning=FALSE, echo=FALSE}
library(ggplot2)
library(ggrepel)
library(tidyverse)
library(gapminder)
library(lubridate)
library(colorspace)
library(viridisLite)
library(viridis)
library(RColorBrewer)
library(sf)
library(hms)
library(rnaturalearth)
library(ggpubr)
library(gganimate)
library(gifski)
```

```{r, echo=FALSE}
usage<-read_tsv("tus_00age_tabular.tsv", na=c(":",":U"))
#view(usage)
names(usage)

```

```{r, echo=FALSE}
timeusageunlabeled <- usage %>%
  pivot_longer("2000":"2010", names_to="year", values_to="Time") %>% separate("freq,unit,sex,age,acl00,geo\\TIME_PERIOD", into=c("frequency","unit","sex","age","activities","state"), sep=",") %>% filter(unit!="PTP_RT") %>% pivot_wider(names_from = unit, values_from = Time) %>% mutate(across("PTP_TIME":"TIME_SP", parse_hm)) %>% mutate(across("PTP_TIME":"TIME_SP", as.duration)) %>% mutate(year = as.numeric(year))
timeusageunlabeled<- timeusageunlabeled[,-1]

activity_mapping<- c(
  "AC0" = "Personal care",
  "AC01" = "Sleep",
  "AC02" = "Eating",
  "AC03" = "Other and/or unspecified personal care",
  "AC1A" = "Main and second job and related travel",
  "AC1B" = "Activities related to employment and unspecified employment",
  "AC1_TR" = "Employment, related activities \nand travel as part of during main and second job",
  "AC2" = "Study",
  "AC212" = "Homework",
  "AC21A" = "School and university",
  "AC22" = "Free time study",
  "AC3" = "Household and family care",
  "AC313" = "Dish washing",
  "AC31A" = "Food management except dish washing",
  "AC321" = "Cleaning dwelling",
  "AC32A" = "Household upkeep except cleaning dwelling",
  "AC331" = "Laundry",
  "AC332" = "Ironing",
  "AC33A" = "Handicraft and producing textiles and other care for textiles",
  "AC342" = "Tending domestic animals",
  "AC343" = "Caring for pets",
  "AC344" = "Walking the dog",
  "AC34A" = "Gardening; other pet care",
  "AC35" = "Construction and repairs",
  "AC36" = "Shopping and services",
  "AC37_39" = "Household management and help family member",
  "AC38A" = "Childcare, except teaching, reading and talking",
  "AC38B" = "Teaching, reading and talking with child",
  "AC4-8" = "Leisure,\nsocial and associative life",
  "AC4-8NSP" = "Unspecified leisure",
  "AC41" = "Organisational work",
  "AC42" = "Informal help to other households",
  "AC43" = "Participatory activities",
  "AC51A" = "Visiting and feasts",
  "AC51B" = "Other social life",
  "AC52" = "Entertainment and culture",
  "AC53" = "Resting",
  "AC611" = "Walking and hiking",
  "AC6A" = "Sports and outdoor activities(not walking and hiking)",
  "AC733" = "Computer games",
  "AC7A" = "Computing",
  "AC7B" = "Hobbies and games(not computing and computer games)",
  "AC811" = "Reading except books",
  "AC812" = "Reading books",
  "AC82" = "TV and video",
  "AC83" = "Radio and music",
  "AC90NSP" = "Unspecified travel",
  "AC913" = "Travel to/from work",
  "AC936" = "Travel related to shopping and services",
  "AC938" = "Transporting a child",
  "AC99NSP" = "Unspecified time use",
  "AC9A" = "Travel except travel related to jobs",
  "AC9B" = "Travel related to study",
  "AC9C" = "Travel related to other household purposes",
  "AC9D" = "Travel related to leisure, social and associative life",
  "TOTAL"="Total")

country_mapping <- c(
  "BE" = "Belgium",
  "BG" = "Bulgaria",
  "DE" = "Germany",
  "EE" = "Estonia",
  "EL" = "Greece",
  "ES" = "Spain",
  "FR" = "France",
  "IT" = "Italy",
  "LV" = "Latvia",
  "LT" = "Lithuania",
  "LU" = "Luxembourg",
  "HU" = "Hungary",
  "NL" = "Netherlands",
  "AT" = "Austria",
  "PL" = "Poland",
  "RO" = "Romania",
  "SI" = "Slovenia",
  "FI" = "Finland",
  "NO" = "Norway",
  "UK" = "United Kingdom",
  "RS" = "Serbia",
  "TR" = "Türkiye"
)
timeusage <- timeusageunlabeled %>%
  mutate(activities_labeled = factor(activities, levels = names(activity_mapping),labels = activity_mapping)) %>% mutate(state_labeled=factor(state,levels=names(country_mapping),labels=country_mapping)) %>% filter(!is.na(TIME_SP))

print(head(timeusage))
```

## HOW PEOPLE SPEND THEIR TIME?

In the surveys, the activities were divided in 8 macroareas: personal care, eat, sleep, work, study, household and family care, leisure, social time, unspecified use of time. In the plot below we can see that the majority of the time is spent on personal care in all european countries. The most vain state is Lithuania(EL), which is also the country with higher sleeping time. On the contrary, in Norway people sleep less.

```{r fig.height=14, fig.width=20, echo=FALSE}
mypalette <-
  c(
    '#88CCEE',
    '#CC6677',
    '#DDCC77',
    '#117733',
    '#888888',
    '#AA4499',
    '#44AA99',
    '#332288'
  )
mainactivities <-
  c("AC0", "AC01", "AC02", "AC1_TR", "AC2", "AC3", "AC4-8", "AC99NSP")
filtereddf <- timeusage %>% 
  filter(age == "TOTAL",
         sex == "T",
         year == 2010,
         activities %in% mainactivities) %>% 
  mutate(my_label = ifelse( TIME_SP> 3600, paste(sprintf("%dH %dmin", TIME_SP %/% 3600, (TIME_SP %% 3600) %/% 60)), " "))

p <-
  timeusage %>% filter(age == "TOTAL",
                       sex == "T",
                       year == 2010,
                       activities %in% mainactivities) %>%ggplot(aes(x = TIME_SP , y = state, fill = activities_labeled))
graph <- p + geom_col(position = 'fill', color = '#333333') + theme_classic() + labs(
  title = 'How people in Europe spend their time?',
  x = 'Percentage of Time Spent on each activity',
  y = 'European country',
  caption = "Source: EUROSTAT, Year:2010"
) + scale_fill_manual(name = "Activities", values = mypalette)  + geom_text(
  aes(label = filtereddf$my_label
  ), 
  position = position_fill(vjust = 0.5),
  color = "black",
  size = 4.5,
  hjust = 0.5
) + theme(plot.title = element_text(
  vjust = 2.0,
  hjust = 0.5,
  color = '#333333',
  size = rel(2.5),
),
  axis.title = element_text(size=rel(2.4)),
  axis.title.x = element_text(margin = margin(t=15)),
  axis.title.y = element_text(vjust = 2),
  axis.text = element_text(size=rel(1.7)),
  legend.position = 'bottom', legend.direction="horizontal", legend.title = element_text(size=rel(2), face="bold"), legend.text = element_text(size = rel(1.25)), legend.box.background = element_rect(),
  legend.box.margin = margin(6, 6, 6, 6)
)

graph
```

Next, we are going to see what are the activities classified as leisure and social life. From the plot, we can see that the leisure activity with the higher time spent is watching television and\\or videos. We find resting and visiting\\feasts respectively in 3rd and 4th place.

```{r fig.width=13}
leisureactivities<-c('AC41','AC42','AC43','AC51A','AC51B','AC52','AC53','AC611','AC6A','AC733','AC7A','AC7B','AC812','AC811','AC82','AC83','AC4-8NSP')

p<- timeusage %>% filter(age=="TOTAL", sex=="T", year==2010, activities %in% leisureactivities) %>%  ggplot(aes(x =reorder(activities_labeled,TIME_SP), y=TIME_SP)) 
p + geom_col(fill='#44AA99') + labs(title = "Different Types of Leisure Activities", x="Leisure Activities", y='Time spent on the activity', caption = "Source: EUROSTAT, Year:2010")+ scale_y_time() + coord_flip() + theme_classic() + theme(
  plot.title = element_text( vjust = 2.0, hjust = 0.5, color = '#333333',
  size = rel(3)),
  axis.title = element_text(size=rel(1.7)),
  axis.title.x = element_text(margin = margin(t=15)),
  axis.text.x=element_text(size=rel(1.2)),
  axis.title.y = element_text(vjust=3)
)
```

In the plot below, I tried to see if there was a correlation between Leisure time and working time. We can see actually that there's a negative correlation. This means that we more time is spent on working, the less is spent in leisure activities. Indeed, the Pearson's correlation coefficient is negative (-0,47). But from the plot we can also see that the country that works the most is Romania, and on the other hand people from Finland spend more than 6 hours in leisure activities. They had the highest amount of leisure time in 2010.

```{r, echo=FALSE}
#scatterplot correlation work and leisure based on the country
leisuretime <- timeusage %>%
  filter(age=="TOTAL", sex=="T", year==2010,activities == "AC4-8") %>%select(TIME_SP,state)
#leisuretime is x
worktime <-timeusage %>% filter(age=="TOTAL", sex=="T", year==2010,activities == "AC01") %>%select(TIME_SP,state)

workingleisure <- merge(leisuretime,worktime, by="state")
head(workingleisure)

p<-workingleisure %>% ggplot(aes(x=TIME_SP.x, y=TIME_SP.y))
p+geom_point(size = 3, alpha = 0.7, color="#117733") + stat_cor(method = "pearson", label.x.npc = 0.75, label.y.npc = 'top')+geom_smooth(method=lm , color="tomato", se=FALSE) + geom_text_repel(label=workingleisure$state, color="#093c1a") + scale_x_time()  + scale_y_time()+ theme_classic() + labs(x="Leisure Time", y="Working Time", title="Leisure time vs Working time", caption="Source: EUROSTAT, Year: 2010") + theme(
  plot.title = element_text( vjust = 2, hjust = 0.5, color = '#333333',
  size = rel(2), face = "bold"),
  axis.title = element_text(size=rel(1.2)),
  axis.title.x = element_text(margin = margin(t=10)),
  axis.text.x=element_text(size=rel(1)),
  axis.title.y = element_text(vjust = 3)
)
```

Now, I wanted to see if there was also a similar linear correlation between studying time and leisure. In this case, there's almost no linear correlation as it is also confirmed by the Pearson's correlation coefficient. Indeed, when this coefficient is near 0, it means that there's no linear correlation.

```{r, echo=FALSE}
normalize <- function(x) {
    (x - min(x)) / (max(x) - min(x))
  }
leisuretime <- timeusage %>%
  filter(age=="TOTAL", sex=="T", year==2010,activities == "AC4-8") %>%select(TIME_SP,state)
#leisuretime is x
studytime <-timeusage %>% filter(age=="TOTAL", sex=="T", year==2010,activities == "AC2") %>%select(TIME_SP,state)

studyingleisure <- merge(leisuretime,studytime, by="state")
#normalized data, but it doesn't change much
studyingleisure_standardized <- studyingleisure %>% mutate(TIME_SP.x = normalize(TIME_SP.x), TIME_SP.y = normalize(TIME_SP.y))

p<-studyingleisure %>% ggplot(aes(x=TIME_SP.x, y=TIME_SP.y))
p+geom_point(size = 3, alpha = 0.7, color='#03396c') + stat_cor(method = "pearson", label.x.npc = 0, label.y.npc = 'top')+geom_smooth(method=lm , color="tomato", se=FALSE) + geom_text_repel(label=workingleisure$state, color='#011f4b') + scale_x_time() + scale_y_time() + labs(x="Leisure Time", y="Studying Time", title="Leisure time vs Studying time", caption="Source: EUROSTAT, Year:2010")  + theme_classic() +theme(
  plot.title = element_text( vjust = 2, hjust = 0.5, color = '#333333',
  size = rel(2), face = "bold"),
  axis.title = element_text(size=rel(1.2)),
  axis.title.x = element_text(margin = margin(t=10)),
  axis.text.x=element_text(size=rel(1)),
  axis.title.y = element_text(vjust = 3)
)
```

From the plot above we can see the Belgium is the country where people study the most. This is confirmed also by the map below. In this map, it's shown the time dedicated to study activities in the countries that participated in the 2010 HETUS. Humorously , a neighbor of Belgium has the lower time spent in studying. This country is Germany, where people spent less than 20 minutes per day to study activities.

```{r, echo =FALSE}
# Maps study ac2 generale
europe <- ne_countries(
  continent = "Europe",
  scale = "large",
  returnclass = "sf"
  )
studytimecountry<- timeusage %>% filter(activities=="AC2", age== "TOTAL", sex=="T", year==2010)%>% select(TIME_SP,state_labeled)
cropped <- st_crop(
  europe, 
  xmin=-12,xmax=34,ymin=34,ymax=71
)
mapsstudy <- inner_join(cropped, studytimecountry, by = c(name = "state_labeled"))
ggplot(data = mapsstudy) +
   geom_sf(data = cropped) +
   geom_sf(aes(fill = TIME_SP/60))+ scale_fill_continuous(type = "viridis", labels= c('20min','25min','30min','35min')) + coord_sf(crs="+proj=aea +lat_1=29.5 +lat_2=42.5") + theme_classic() +labs(title="Studying time in Europe", fill="Studying time",caption= str_wrap("In the map, only the countries that participated in the survey are shown. Source:EUROSTAT", width = 50), subtitle='in 2010')+theme(
     plot.title = element_text( vjust = 1.5, hjust = 0.5, color = '#333333',
  size = rel(2), face = "bold"),
  plot.subtitle = element_text(hjust=0.9),
  plot.caption = element_text(hjust=1),
  legend.title = element_text(size = rel(1), margin = margin(b=10), face = 'bold'),
  legend.title.align=0.5,
  legend.box.background = element_rect(),
  legend.box.margin = margin(4,4,4,4),
  legend.text = element_text(size = rel(0.75))

  )
  
```

In the following plot, I wanted to investigate why Belgium had more studying time than the others. Belgium is highlighted with the color red. As we can see, Belgium people spend a lot of time in school or university and in doing homework.

```{r, echo=FALSE}
studyactivities<-c("AC212","AC21A","AC22")
timeusage %>% filter(age=="TOTAL", sex=="T", activities %in% studyactivities, year==2010) %>% ggplot(aes(x =TIME_SP, y =state)) + geom_col(fill='#888888',show.legend = FALSE) + geom_col(fill='tomato', data=timeusage%>% filter(age=="TOTAL", sex=="T", activities %in% studyactivities, year==2010, state=='BE'), show.legend = FALSE) + facet_wrap(~activities_labeled)+ scale_x_time()  + labs(title='Comparison between European countries in studying activities', caption='Source: EUROSTAT, Year: 2010', x="Time Spent") + theme_classic() + theme(
  plot.title = element_text(size = rel(1.5),hjust = 0.5, vjust = 2),
  axis.text.x = element_text(angle=60,hjust=1),
  axis.title.y=element_blank(),
  axis.title.x = element_text(size=rel(1.2)),
  panel.grid = element_blank(),
  strip.text = element_text(face='bold'),
  panel.background = element_rect(color='black')
  #strip.background = element_rect(fill='#FFDCD6')
)
```

In the next plot, I wanted to see the difference time spent on each main activities between females and males. There are no big differences except for time dedicated to work and household and family care.

```{r fig.width=12, echo=FALSE}
#How men and Women spend their time?
 timeusage %>% filter(age=="TOTAL", sex!="T", activities %in% mainactivities, year==2010) %>% ggplot(aes(x =sex , y =TIME_SP, fill=sex)) + geom_col(show.legend = FALSE)+  facet_wrap(~activities_labeled) + theme_classic() + scale_y_time()+ labs(title = 'How Men and Women spend their time?', y='Time Spent', caption = "Source: EUROSTAT, Year:2010") + theme(plot.title =  element_text(
  vjust = 2.0,
  hjust = 0.5,
  color = '#333333',
  size = rel(3),
  margin = margin(b=12)
),
  axis.title = element_text(size=rel(1)),
  axis.title.x = element_blank(),
  axis.title.y = element_text(size=rel(1.25), vjust=2.7),
  axis.text = element_text(size=rel(0.75)),
  panel.grid = element_blank(),
  panel.background = element_rect(color='black'),
  strip.text = element_text(margin = margin(4,4,4,4), size= 10, face='bold')
  
)
```

In the plots below, we can see the difference in time spent in household and family care. Female tend to spend more time than males. We can see this phenomenon both in the year 2000 and 2010.

```{r, echo=FALSE}
p<- timeusage %>% filter(age=="TOTAL", sex!="T",activities == "AC3") %>% ggplot(aes(x=sex, y=TIME_SP, color=sex, fill=sex)) + geom_violin(draw_quantiles = c(.25,.5,.75), alpha=0.3, show.legend = FALSE) + facet_wrap(~year) + scale_y_time()+ labs(title='Time spent in Household and family care activities in the years 2000 and 2010', subtitle='by Gender', caption='Source:EUROSTAT', y= 'Time Spent') + theme_bw() + theme(plot.title =  element_text(
  vjust = 2.0,
  hjust = 0,
  color = '#333333',
  size = rel(1),
  margin = margin(b=4),
  face='bold'
),
  plot.subtitle = element_text(size=rel(1.1), margin=margin(b=8), hjust=1),
  axis.title = element_text(size=rel(0.75)),
  axis.title.x = element_blank(),
  axis.text.x = element_text(size=11),
  axis.title.y = element_text(vjust=2.5 ,size=rel(1.75)),
  axis.text.y =  element_text(size=rel(0.9)),
  panel.grid = element_line(color = '#ededed'),
  panel.grid.major.x = element_blank(),
  strip.text = element_text(margin = margin(6,6,6,6), size= rel(1.1), color='white', face='bold'),
  strip.background = element_rect(fill = '#555555')
  
)
p
```
